---
title: "20210112 National Soils Data Processing Final Script"
author: "Rachelle LaCroix"
date: "1/12/2021"
output: github_document
---

# This script imports and wrangles all national soils data files into single dataframe.
### Author: Itamar Shabtai - modified by R. LaCroix
#### Version: 2021-01-12  
  
Imports the following data streams:
  + pH and carbonates  
  + size fractions  
  + moisture data  
  + TOC  
  + HH extractions  
  + CEC  
  + particle size  
  
Calculates and normalizes for final data points and writes csv files for each dataset.  
Then all data sets are joined together using `pedon_ID` and `horizon` with the master metadata file (updated in December 2020) to produce one final data frame (`final_data`) to export for use. **The resulting dataframe contains all horizons from all pedons sampled.**   Missing data points are represnted with `NA` All final variable descriptions will be added to final excel document.  
```{r setup}
knitr::opts_chunk$set(echo = TRUE)
# Libraries

library(tidyverse)
library(here)
library(readxl)
```

```{r}
# Import data files----

pH_carbonate <- 
  read_csv("./raw_data/pH and carbonate/pH_carbonate_data_summary_final.csv",
           col_types = cols(horizon = col_factor(levels = c("a","c","o"))))

size_fractions <- 
  read_csv("./raw_data/Size fractions/size_fractions_raw_data.csv",
           col_types = cols(horizon = col_factor(levels = c("a","c", "e"))))

moisture <- 
  read_csv("./raw_data/Moisture/soil_moisture_raw_data.csv", 
           col_types = cols(horizon = col_factor(levels = c("a","b", "c", "e", "o"))))

missing_HWC <- 
  read_csv("./raw_data/Moisture/missing_HWC.csv",
           col_types = cols(horizon = col_factor(levels = c("a","b", "c", "e", "o")))) %>% 
  select(1, 5) 

TOC_final_summary <- 
  read_csv("./raw_data/Microbial biomass/TOC_final_summary.csv",
           col_types = cols(horizon = col_factor(levels = c("a","b", "c", "e", "o"))))

HH_extraction <- 
  read_csv("./raw_data/HH extraction/HH_extraction_raw_plusinfo.csv",
           col_types = cols(horizon = col_factor(levels = c("a","c","e"))))

bulk_soil_CN_isotopes <- 
  read_csv("./raw_data/Elemental and isotopes/bulk_soil_CN_natu_abund_raw_data.csv",
           col_types = cols(horizon = col_factor(levels = c("a","c","e"))))

all_pedon_meta_data <- 
  read_xlsx("./raw_data/master_pedon_metadata.xlsx")

CEC <- 
  read_csv("./raw_data/CEC/CEC_data.csv", 
           col_types = cols(horizon = col_factor(levels = c("a","b", "c", "e", "o")))) %>% 
  select(-1) 

particle_size <- 
  read_csv("./raw_data/Particle size distribution/size_data.csv", 
                          col_types = cols(horizon = col_factor(levels = c("a","c"))))

# Tidy moisture and bulk density----

# calculating gravimetric (GWC), volumetric (VWC), hygroscopic (HWC) water contents, bulk density (db)

core_volume <- 79.52156404 # volume of bulk density cores (cm^3)

moisture_calculated_values <- moisture %>% 
  dplyr::select(3:5,7,9,10,11,12,13) %>% # selecting relevant columns
  #drop_na(bag_wet_soil) %>% # removing missing samples
  #drop_na(tin_airdry_weight) %>% 
  mutate(wet_weight = bag_wet_soil - bag_weight,
         ovendry_weight = tin_ovendry_weight - tin_weight,
         GWC = (wet_weight - ovendry_weight)/ovendry_weight,
         db = ovendry_weight/core_volume,
         VWC = GWC*db,
         HWC = (tin_airdry_weight - tin_ovendry_weight)/ovendry_weight) %>% 
  group_by(pedon_ID, horizon, location) %>% # calculating means from replicates
  summarise(mean_GWC = mean(GWC, na.rm = TRUE),
            mean_HWC = mean(HWC, na.rm = TRUE),
            mean_VWC = mean(VWC, na.rm = TRUE),
            mean_db = mean(db, na.rm = TRUE
            )) 

#write_rds(x = moisture_calculated_values, file = here("data/moisture_bulk_density.rds"))
#write_csv(moisture_calculated_values, "moisture_bulk_density.csv")

# Tidy pH and carbonate----

# all negative inorganic C assumed equal to zero

# convert inorganic C content to %carbonate in soils

mw_carbonate <- 100.0869 # molar mass of calcium carbonate

mw_c <- 12.0107 # molar mass of C

pH_carbonate_percent <- pH_carbonate %>% 
  mutate(percent_carbonate = mg_IC_g_soil_zero/1000*(mw_carbonate/mw_c)*100)

#write_rds(x = pH_carbonate_percent, file = here("data/pH_carbonate.rds"))
#write_csv(pH_carbonate_percent, "pH_carbonate.csv")

# Tidy size fractions----

# calculate fraction amount in g, relative amount and recovery

# silt and clay pan weight were added in excel to avoid NA's

size_fractions_calculated_values <- size_fractions %>% 
  mutate(coarse_fraction_g = sand_and_POM_and_pan_weight_g - sand_pom_pan_weight_g,
         fine_fraction_g = total_sc_and_pan_g - total_sc_pan_g,
         recovery_percent = (coarse_fraction_g + fine_fraction_g) / soil_weight_g * 100,
         relative_coarse_fraction = coarse_fraction_g / soil_weight_g,
         relative_fine_fraction = fine_fraction_g / soil_weight_g,
         fine_coarse_ratio = relative_fine_fraction / relative_coarse_fraction)

#write_rds(x = size_fractions_calculated_values, file = here("data/size_fractions.rds"))
#write_csv(size_fractions_calculated_values, "size_fractions.csv")

# Tidy HH Extraction----

# calculate concentrations of HH-extractable C, Fe and Al per g soil

HH_volume <- 0.01 # extracting solution was 10 mL

HH_dilution <- 10 # all ICP samples diluted 1:9

HH_calculated_values <- HH_extraction %>% 
  mutate(HH_DOC_ppm = npoc_ppm * toc_dilution,
         HH_DOC_mg_g = HH_DOC_ppm * HH_volume / soil_weight_g,
         Fe_mg_g = Fe_ppm * HH_dilution * HH_volume / soil_weight_g,
         Al_mg_g = Al_ppm * HH_dilution * HH_volume / soil_weight_g,
         Fe_plus_Al = Fe_mg_g + Al_mg_g)

#write_rds(x = HH_calculated_values, file = here("data/HH_extraction.rds"))
#write_csv(HH_calculated_values, "HH_extraction.csv")

# Tidy SOC---- 

# Subtract IC from TC to find OC

calculated_bulk_CN <- left_join(bulk_soil_CN_isotopes, pH_carbonate_percent, by = c("pedon_ID", "horizon", "rep"))  

calculated_bulk_CN[c("mg_IC_g_soil_zero", "mg_IC_g_soil")][is.na(calculated_bulk_CN[c("mg_IC_g_soil_zero", "mg_IC_g_soil")])] <- 0

calculated_bulk_CN <- calculated_bulk_CN %>% 
  mutate(SOC = percent_C - mg_IC_g_soil_zero*0.1) %>% 
  mutate(TN = percent_N)  

#write_rds(x = calculated_bulk_CN, file = here("data/bulk_CN.rds"))
#write_csv(calculated_bulk_CN, "bulk_CN.csv")

# Tidy microbial biomass C----

NPOC_volume <- 0.05 # extraction volume equals 50 mL

calculated_MBC_DOC <- TOC_final_summary %>% 
  mutate(DOC_mg_g_wet = NPOC * NPOC_volume / unfum_weight,
         DN_mg_g_wet = TN * NPOC_volume / unfum_weight,
         MBC_mg_g_wet = fum_NPOC * NPOC_volume / fum_weight,
         MBN_mg_g_wet = fum_TN * NPOC_volume / fum_weight) %>% 
  group_by(pedon_ID, horizon) %>% 
  summarize(mean_DOC_wet = mean(DOC_mg_g_wet),
            mean_DN_wet = mean(DN_mg_g_wet),
            mean_MBC_wet = mean(MBC_mg_g_wet),
            mean_MBN_wet = mean(MBN_mg_g_wet))

#write_rds(x = calculated_MBC_DOC, file = here("data/MBC_DOC.rds"))
#write_csv(calculated_MBC_DOC, "MBC_DOC.csv")

# Tidy CEC----

#joins moisture df for unique_ID

CEC_tidy <- CEC %>% 
  left_join(moisture_calculated_values, by = c("horizon", "pedon_ID"))%>%
  select(-c(mean_GWC, mean_HWC, mean_VWC, mean_db)) %>% 
  rename(CEC = Sum) 

#write_rds(x = CEC_tidy, file = here("data/CEC.rds"))
#write_csv(CEC_tidy, "CEC.csv")

# Tidy mechanical composition----

#joins moisture df for unique_ID

particle_size_tidy <- particle_size %>% 
  left_join(moisture_calculated_values, by = c("horizon", "pedon_ID"))%>%
  select(-c(sample_num,mean_GWC, mean_HWC, mean_VWC, mean_db)) 

#write_rds(x = particle_size_tidy, file = here("data/mechanical_composition.rds"))
#write_csv(particle_size_tidy, "mechanical_composition.csv")

# Only relevant variable and unique_ID in each dataframe----

ph_carbonate_short <- dplyr::select(pH_carbonate_percent, 2,3, 6, 8)
CN_short <- dplyr::select(calculated_bulk_CN, 2,3,5, 6, 15,16)
HH_short <- dplyr::select(HH_calculated_values,  14,15, 16:20)
size_short <- dplyr::select(size_fractions_calculated_values, 2,3, 17, 18,19)
CEC_short <- dplyr::select(CEC_tidy,  1,2,3,5, 6, 7, 8, 9)
particle_short <- dplyr::select(particle_size_tidy,1, 2,3, 4, 5, 6, 7, 8)


# joining data frames by unique_ID----

# moisture dataframe has horizons

joined_data <- left_join(all_pedon_meta_data, moisture_calculated_values, by = c("horizon", "pedon_ID"))%>%
  select(-location.y)%>%  # drop column because names of sites are spelled differently and creates extra column
  rename(location = location.x)
joined_data <- left_join(joined_data, ph_carbonate_short,  by = c("horizon", "pedon_ID")) 
joined_data <- left_join(joined_data, CN_short,  by = c("horizon", "pedon_ID")) 
joined_data <- left_join(joined_data, HH_short,  by = c("horizon", "pedon_ID")) 
joined_data <- left_join(joined_data, size_short,  by = c("horizon", "pedon_ID"))
joined_data <- left_join(joined_data, CEC_short,  by = c("horizon", "pedon_ID"))%>%
  select(-location.y)%>%  # drop column because names of sites are spelled differently and creates extra column
  rename(location = location.x)
joined_data <- left_join(joined_data, calculated_MBC_DOC,  by = c("horizon", "pedon_ID"))
joined_data <- left_join(joined_data, particle_short,  by = c("horizon", "pedon_ID")) %>% 
  dplyr::select(pedon_ID, horizon, everything())%>%
  select(-location.y)%>%  # drop column because names of sites are spelled differently and creates extra column
  rename(location = location.x)

# normalizing variable to oven dry weight----

joined_data_moisture_corrected <- joined_data %>% 
  #filter(horizon == c("a", "c")) %>% 
  mutate(DOC = mean_DOC_wet * (mean_GWC + 1),
         DN = mean_DN_wet * (mean_GWC + 1),
         MBC = mean_MBC_wet * (mean_GWC + 1),
         MBN = mean_MBN_wet * (mean_GWC + 1),
         SOC = SOC * (mean_HWC + 1),
         TN = TN * (mean_HWC + 1),
         IC = percent_carbonate * (mean_HWC + 1),
         HH_DOC = HH_DOC_mg_g * (mean_HWC + 1),
         Fe = Fe_mg_g * (mean_HWC + 1),
         Al = Al_mg_g * (mean_HWC + 1),
         Fe_Al = Fe_plus_Al * (mean_HWC + 1))

# normalizing to unit OC and TN----

final_data <- joined_data_moisture_corrected %>% 
  mutate(MBC_C = MBC / (SOC / 100),
         MBN_N = MBN / (TN / 100),
         DOC_C = DOC / (SOC / 100),
         DN_N = DN / (TN / 100),
         HH_DOC_C = HH_DOC / (SOC / 100)) 

#write_rds(x = final_data, file = here("data/normalized_final_data.rds"))
#write_csv(final_data, "final_data_national_soils.csv")
```
